---
title: "STAT 331 Group Project"
author: "Katrina, Matina, Lindsey, Kaviya"
format: 
  html:
    self-contained: true
    embed-resources: true
    code-tools: true
    toc: true
    code-fold: true
editor: visual
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

## Our Hypothesis

Higher CO2 emissions per capita are associated with lower life expectancy.

## About the Data

The data consists of two datasets that come from [Gapminder](https://www.gapminder.org/data/): Life Expectancy (`life_expectancy_at_birth.csv`) and Consumption-Based CO2 Emissions Per Capita (`co2_pcap_cons.csv`).

In the dataset for Life Expectancy (`life_expectancy_at_birth.csv`), the data collected includes the number of years a newborn infant would live if the current mortality rates at different ages were to stay the same throughout its life across different countries. The dataset originally had 195 rows and 302 columns. After filtering and cleaning, the data consists of 186 rows and 24 columns with mainly double data types. Each row represents a country, with each column representing the life expectancy for a particular country across years from 2000 to 2022. More information about this dataset and its documentation can be found here: [Life Expectancy](https://www.gapminder.org/data/documentation/gd004/).

In the dataset for Consumption-Based CO2 Emissions Per Capita (`co2_pcap_cons.csv`), the data collected includes the average amount of carbon dioxide emissions per person, taking into account consumption patterns, across different countries. The dataset originally had 194 rows and 224 columns. After filtering and cleaning, the data consists of 194 rows and 24 columns with mainly double data types. Each row provides information about the CO2 emissions per capita for a particular country across years from 2000 to 2022. More information about this dataset and its documentation can be found here: [CO2 Emissions Consumption Historic Dataset](https://docs.google.com/spreadsheets/d/1t_GOXGrXQgSasIB7dWfRrq9dkqjrqwy3xHvywH7ah1A/edit#gid=501532268).

## Data Set-Up

```{r}
# Accessing libraries
library(tidyverse)
library(janitor)
library(moderndive)
library(patchwork)
```

```{r}
# Load in the life expectancy csv
life_exp <-read_csv(here::here("Datasets/life_expectancy_at_birth.csv"))

#load in the carbon emissions data 
carbon <-read_csv(here::here("Datasets/co2_pcap_cons.csv"))

```

## Data Cleaning

```{r}
#cleaning carbon
clean_carbon <- carbon %>% 
  na.omit() %>% 
  mutate(across(.cols = `2000`:`2022`, 
                .fns = ~as.numeric(.x))) %>% 
  select(country, `2000`:`2022`)

clean_life_exp <- life_exp %>% 
  na.omit() %>% 
  mutate(across(.cols = `2000`:`2022`, 
                .fns = ~as.numeric(.x))) %>% 
  select(country, `2000`:`2022`)
```

## Pivot the Data Longer

```{r}
# Pivot longer the clean carbon data: 
clean_carbon <- clean_carbon %>%
  pivot_longer(cols = !country,
               names_to = "year",
               values_to = "carbon_emissions_tonnes")


# Pivot longer the clean life expectancy data:
clean_life_exp <- clean_life_exp %>%
  pivot_longer(cols = !country,
               names_to = "year",
               values_to = "life_expectancy_years")
```

## Joining the Data Sets:

```{r}
#joining the two cleaned data sets 
joined_data <- left_join(clean_carbon, clean_life_exp, by = c("country", "year"))

```

# Linear Regression

### Visual 1 Table

```{r}
#The relationship between the two quantitative variables changing over time. 

av_data <- joined_data %>% 
  group_by(country) %>% 
  summarise(avg_crbn = mean(carbon_emissions_tonnes, na.rm=T),
         avg_life = mean(life_expectancy_years, na.rm=T))


```

### Visual 2 Table

```{r}
#finding the countries with the top 10 carbon emissions over 22 year average
top_ten_carbon <- joined_data %>% 
  group_by(country) %>% 
  summarise(avg_crbn = mean(carbon_emissions_tonnes, na.rm=T),
         avg_life = mean(life_expectancy_years, na.rm=T)) %>%
  #slicing the countries with the top 10 carbon emissions so we can see what they are and how the life expectance vs carbon emissions changes over time
  slice_max(avg_crbn, n=10) %>% 
  arrange(desc(avg_crbn))

```

# Data Visuals

### Data Visual 1

Below we created a visualization that illustrates the relationships between consumption-based C02 emissions per capita and life expectancy over the years 2000 to 2022, divided into three decades. Our response variable (y) is life expectancy (in years) and our explanatory variable (x) is year (2000-2022). In the 2000's the increase in carbon emissions appear to have a larger impact on life expectancy compared to the 2010's and 2020's. In the 2000's and the 2020's, countries with larger carbon emissions had a lower life expectancy while in the 2010's the life expectancy continued to slightly increase. In all three decades, the life expectancy began to plateau at around 15 tonnes of carbon emission.

```{r}
joined_data %>% 
  mutate(year=as_factor(year),
          year=fct_collapse(.f=year,
                           "2000's"= 2000:2009,
                            "2010's"= 2010:2019,
                            "2020's"=2020:2022)) %>% 
  ggplot(mapping = aes (x = carbon_emissions_tonnes,
                        y = life_expectancy_years)) +
  facet_grid(~ year) +
  geom_smooth(method = "loess")+
  geom_jitter(alpha = 0.1) + 
  scale_color_brewer(palette = "Dark2") +
  labs(x = "CO2 Emissions Per Capita (tonnes)",
       subtitle = "Life Expectancy",
       y = "") +
  theme_bw()

```

### Data Visualization 2

Below we have created another visualization to show the relationship between the consumption-based C02 emissions per capita and the life expectancy averaged per country over the years 2000 to 2022. The graph indicates that there is likely not a linear relationship between the two variables. As the average carbon emissions increase between 0 and 5 tons, there is a sharp increase in average life expectancy. However life expectancy begins to plateau to around 75-80 years at 10 to 15 tons of carbon emissions. The overall trend is that as carbon emissions increase, life expectancy also increases, though at different rates.

```{r}
joined_data %>% 
  group_by(country) %>% 
  summarise(avg_crbn = mean(carbon_emissions_tonnes, na.rm=T),
         avg_life = mean(life_expectancy_years, na.rm=T)) %>% 
  ggplot(mapping = aes(x = avg_crbn,
                       y = avg_life)) +
  geom_jitter(alpha = 0.2) +
  geom_smooth(method = "loess") +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "Average Carbon Emissions per Capita (tonnes)",
       subtitle = "Average Life Expectancy",
       y = "")
  
```

# Linear Regression

We used linear regression to model the relationship between life expectancy (response variable) and carbon emissions (explanatory variable).

```{r}
library(broom)

#create a new dataset similar to the top 10 country one but Instead let's use the data from all the countries to see the correlation between a countries average carbon emissions with the average life expectancy 
lm_data <- joined_data |>
  group_by(country) |>
  summarise(avg_crbn = mean(carbon_emissions_tonnes, na.rm=T),
         avg_life = mean(life_expectancy_years, na.rm=T))

lm_crbn_life <- lm(avg_life~avg_crbn, data = lm_data)
lm_crbn_life
```

The model represents the relationship using the equation below:

$$AverageLifeExpectancy = \beta_0 + \beta_1 * AverageCarbonEmissions + ε$$

B0: This coefficient represents the average life expectancy when the average carbon emissions equal zero.

B1: This coefficient represents the average carbon emissions and how much the average life expectancy changes for each additional unit of carbon emissions.

ε: This represents the error term, which is the difference between the actual observed life expectancy and predicted life expectancy.

## Regression Model

The fanning of residuals suggests that a linear model is not appropriate and that equal variance is non-normal.

```{r}
lm_crbn_life %>% 
  ggplot(aes( x = .fitted , y=.resid))+
  geom_jitter()+
  geom_smooth(method = "loess")
```

## Model Fit

Using the values from the regression model, the table is constructed below with the following variables: `var_y`, `var_y_hat`, `var_residual`, and `r_squared`. `var_y` represents the variance in the response values. `var_y_hat` represents the variance in the fitted values. `var_residual` represents the variance in the residuals. `r_squared`represents the proportion of the spread/variation of the outcome variable `avg_life` explained by the model.

To evaluate the quality of the regression model, we compute the R^2^ value. The R^2^ value for the regression model is approximately 0.342. This means that around 34.2% of the variability in the response values is accounted for by the model.

In terms of the "quality" of the model, this suggests a moderate level of fit. While the model does explain some of the variance in the response variable, a significant portion (approximately 65.8%) of the variance is not explained by the model. This could mean that there are other variables not included in the model that are influencing the response, or that the model is not capturing the relationship between the variables effectively.

```{r}
model_fit <- get_regression_points(lm_crbn_life) %>% 
  summarize(var_y = round(var(avg_life),1), 
                      var_y_hat = round(var(avg_life_hat),1), 
                      var_residual = round(var(residual),1),
                      r_squared = round(1-(var_residual/var_y),3))

knitr::kable((model_fit),format = "html")
```

### 3.1

With your simple linear regression, generate predictions using the `predict()` function. Then, add random errors to the predictions, using the residual standard error estimated from the linear regression model (acquired with `sigma()`).

Now, compare these simulated observations to the observed data. Generate the **same** plot of the relationships modeled by the linear regression, for the simulated data (that you made at during [2.1 Data Visualizations](https://earobinson95.github.io/stat331-calpoly/final-project/final-project-details.html#dataviz)).

> Plot the visualization of the observed data and the simulated data side-by-side.
>
> Discuss how the simulated data are similar and / or different to the observed data.

```{r}
predictions <- predict(lm_crbn_life)
sig_data <- sigma(lm_crbn_life) 

noise <- function(x, mean = 0, sd){
  x + rnorm(length(x), 
            mean, 
            sd)
}


simulations <- tibble(predicted_lf = noise(predictions, 
                                           sd = sig_data)
                   )

sim_data <- lm_data |> 
  filter(!is.na(avg_crbn), 
         !is.na(avg_life)
         ) |> 
  select(avg_crbn, avg_life) |> 
  bind_cols(simulations)

head(sim_data)
```

```{r}
predicted_plot <- sim_data %>% 
  ggplot(mapping = aes(x = avg_crbn,
                       y = predicted_lf)) +
  geom_point() +
  labs(x = "Avg Carbon Emissions",
       y = "",
       subtitle = "Predicted Life Expectancy (years)",
       title = "Simulated Life Expectancy based on Regression Model ") +
  theme_bw()

# the dataset here needs to be changed
observed_plot <- lm_data %>%
  ggplot(mapping = aes(x=avg_crbn,
                       y = avg_life)) +
  geom_point() +
  labs(x = "Avg Carbon Emissions",
       y = "",
       subtitle = "Observed Life Expectancy (years)",
       title = "Observed Data") +
  theme_bw()

observed_plot + predicted_plot

  
```

## 3.2

```{r}
nsims <- 1000
sims <- map_dfc(.x = 1:nsims,
                .f = ~ tibble(sim = noise(predictions, 
                                          sd = sig_data)
                              )
                )

head(sims)

```

```{r}
colnames(sims) <- colnames(sims) |> 
  str_replace(pattern = "\\.\\.\\.",
                  replace = "_")

sims
```

```{r}
sims <- lm_data |> 
  filter(!is.na(avg_crbn), 
         !is.na(avg_life)) |> 
  select(avg_life) |> 
  bind_cols(sims)

```

```{r}
sim_r_sq <- sims |> 
  map(~ lm(avg_life ~ .x, data = sims)) |> 
  map(glance) |> 
  map_dbl(~ .x$r.squared)
```
```{r}
head(sim_r_sq)
```
```{r}
sim_r_sq <- sim_r_sq[names(sim_r_sq) != "avg_life"]
```

```{r}
tibble(sims = sim_r_sq) |> 
  ggplot(aes(x = sims)) + 
  geom_histogram(binwidth = 0.025) +
  labs(x = expression("Simulated"~ R^2),
       y = "",
       subtitle = "Number of Simulated Models") +
  theme_bw()
```

